\bf Model Visualization Techniques for a Social Network Model

### 2017-09-19 paper.Rnw
making progress on reorganizing section 4.

%---------------------------------------------------
%                 Placing Figures

@@

\st{
%\par In order to estimate the parameters of  SAOMs the software SIENA, and its \texttt{R} implementation \texttt{RSiena}, was developed \cite{RSiena}. This software marks a huge contribution to the field of social network analysis., and when we began working with it we were eager to understand how it works. HH: we are still interested in this.
%\hh{The next sentence is very vague - make it more specific: what aspect of SAOMs is not well understood? I think what you want is something like ``SAOMs are not very tractable analytically with traditional methods, e.g. likelihood functions turn into very complex objects because of the sheer number of parameters involved. Therefore, computationally more tractable solutions are used to fit estimators, in particular, SAOMs are fit using ... }
Unlike other network models, SAOMs are not very well understood. \st{They are relatively new, especially compared to the classic exponential random graph models, and they are not very tractable analytically. Likelihood functions quickly very complex objects to analyze due to the dependency structure inherent in the data. Therefore, computationally more tractable solutions are used to fit estimators, and in particular, % Traditional modeling steps, like determining the likelihood function, are very complex. Because of this, 
SAOMs are often fit to data using a series of Markov Chain Monte Carlo (MCMC) phases for finding method of moments estimators. In order to estimate the parameters of SAOMs, we use the software SIENA, and its \texttt{R} implementation \texttt{RSiena}, which was developed by \citet{RSiena}. This software marks a huge contribution to the field of social network analysis, but the many moving pieces involved in parameter estimation are largely ``behind the scenes" and hidden from the software user.} In this paper, in order to better understand the model-fitting process, we attempt to bring SAOM fits and the fitting process out of their black boxes, by combining the principals of network visualization with those of model visualization as discussed in \citet{modelvis}. %Here, we explore RSiena and SAOMs with innovative network visualizations, both static and dynamic.
By bringing some light to the underlying methods and structures that are behind the scenes when fitting SAOMs to network data, we aim to help researchers working with the models better understand the implications and analyses of these models.
}

@@

The \texttt{RSiena} software that fits the models to data also includes a goodnes-of-fit function for examining model fit, \texttt{sienaGOF()}. This function ``assess[es] the fit of the model with respect to auxiliary statistics of networks" \citep[p.~53]{RSienamanual}. Examples of auxiliary statistics include the out- or indegree distribution on the nodes, with the option for users to input their own statistics to examine. The goodness-of-fit is evaluated as follows:
\begin{enumerate}
\item The auxiliary statistics are computed on the observed data and on $N$ simulated observations from the model. Typically, $N=1000$. 
\item The mean vector and covariance matrix of the statistics on the simulations from the model are computed.
\item The Mahalanobis distance from the observed statistics to the distribution of the simulated statistics is computed using the mean and covariance found in step 2. 
\item The Mahalanobis distrance from each of the $N$ simulations to the same distribution is computed, and the Mahalanobis distance of the observed data is compared to this distribution of distances.
\item A $p$-value is found by computing the proportion of simulated distances found in step 4 that are as large or larger than the Mahalanobis distance from the data. A SAOM is thus considered a good fit if $p$ is large.
\end{enumerate}

@@

The \texttt{plot.sienaGOF()} function allows us to visualize this fit. %In Figure~\ref{fig:sienagof}, we provide an example of what the goodness-of-fit plot output looks like for indegree and outdegree statistics for a small model. 
This function draws a box plots and a violin plot at each value of the statistic of interest observed in the simulations. In order to compare the distribution of the simulated network to the true data, red points connected by red lines representing the observed values are superimposed on the boxplots. If the red points lie "well within" the simulated values, the model is a good fit to the data. The plot is not intuitive for understanding network data. %The model examined in Figure~\ref{fig:sienagof} appears to do a better job of capturing the outdegree distribution than the indegree distribution of the data. 
In Section~\ref{sec:minds}, we propose a new way of visualizing goodness-of-fit that uses the traditional node-link diagram to visualize the entire network instead of numerical summaries of the the network.

@@

<<alldataviz, fig.cap="The smaller friendship network data we will be modelling throughout the paper. In the first wave, we can see there are two large, separate friend group. By the second wave, three students with heavier drinking behavior have separated from their group, while the others become members of the other group. By the third wave, the large group has almost completely broken up. We want to capture these changes with our SAOMs.", out.width='.99\\textwidth', fig.height=3>>=
actual1 <- merge(data.frame(as.edgelist(as.network(fd2.w1))), 
                 data.frame(id = 1:16, drink = drink2[,1]),
                 by.x = "X1", by.y = "id", all = T)
actual1$Wave <- 1 
actual2 <- merge(data.frame(as.edgelist(as.network(fd2.w2))),
                 data.frame(id = 1:16,drink = drink2[,2]),
                 by.x = "X1", by.y = "id", all = T)
actual2$Wave <- 2
actual3 <- merge(data.frame(as.edgelist(as.network(fd2.w3))),
                 data.frame(id = 1:16,drink = drink2[,3]),
                 by.x = "X1", by.y = "id", all = T)
actual3$Wave <- 3
alldat <- rbind(actual1, actual2, actual3)
#alldat$X1 <- paste0("V", alldat$X1)
alldat$X1 <- as.factor(alldat$X1)
#alldat$X2 <- ifelse(!is.na(alldat$X2), paste0("V", alldat$X2), NA)
alldat$X2 <- as.factor(alldat$X2)
#write_csv(alldat, "smallfriends4Geomnet.csv")
alldataviz <- ggplot(data = alldat, 
       aes(from_id = X1, to_id = X2, color = as.ordered(drink), group = as.ordered(drink))) +
  geom_net(fiteach = FALSE, directed=TRUE, layout.alg = 'fruchtermanreingold', size = 5, arrowsize = .25, arrowgap = .05, ecolour = 'grey40', labelon=T, labelcolour = 'black', fontsize = 4, vjust = .5) +
  scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior", labels = c("never", "1-2 times per year", "once per month", "once per week")) + 
  ThemeNet +
  facet_wrap(~Wave, labeller = "label_both") + 
  theme(legend.position = 'bottom', 
        panel.background = element_rect(fill = NA, color = "grey30"), 
        legend.text = element_text(size = 12, family = "Times New Roman"), 
        legend.title = element_text(size = 14, family = "Times New Roman"))
alldataviz
@

@@

Every good data analysis includes both numerical and visual summaries of the data, so why restrict model description and diagnostics to numerical summaries? The concept of model visualization was developed to complement traditional model diagnostic tools. Typically, numerical summaries such as $R^2$ are used to assess model fit, and the occasional visualization, like a residual plot, are used to determine how well the model fits the data. \citeauthor{modelvis} outline three separate ideas, each of which are can be referred to simply as the "model": the model family, the model form, and the fitted model. The latter is primarily what one thinks of first when considering a model in a data analysis, where specified a model is fit to data, and parameter estimates and other numerical summaries, such as $R^2$ are reported. In the context of SAOMs, the fitted model contains the form of the rate and objective functions, the estimated rate parameters, and the estimated objective function parameters. The model form describes the the model \textit{before} the fitting process, defining which parameters are in the model within the context of the larger model family. In SAOMs, the model form includes description of the rate and objective functions and the variables therein that describe how the network evolves over time. Finally, the model family is the broadest description of the model. This is the type of model that you wish to fit to the data, and is chosen based on the problem, data, and knowledge at hand. For example, we chose to use a SAOM to model network data over an exponential random graph model (ERGM) because we believe that actor-level variables effect network structure and formation, and we wanted to model the network changes over time. These three uses of the term ``model" can each be visualized according to the three principals of model visualization: display the model in the data space, look at a collection of models, and explore the process of fitting the model, not just the end result. We turn our focus to the fitted model and the model form. Specifically, we want to know how the model form affects the fitted model. Using our example data sets and our visualization toolbox, we hope to answer this question for SAOMs.

@@

where $s_{i3} = \sum_{j \neq h} x_{ij}x_{ih}x_{hj} \mathbb{I}(z_i = z_h \neq z_j)$, and $s_{i4} =  |\{j : x_{ij} = 0, \sum\limits_h x_{ih}x_{hj} \geq 2\}|$. These statistics are known as the number of jumping transitive triplets and the number of doubly achieved distances two effect, respectively. The first statistic emphasizes triad relationships that are formed between actors from different covariate groups, while the other emphasizes indirect ties between actors. These two effects are visually represented and further described in Figure~\ref{fig:jtt} and Figure~\ref{fig:dad}, respectively.

@@

\begin{table}
\scalebox{0.9}{
\begin{tabular}{l|ccc|ccc }
  \hline
  & \multicolumn{3}{|c|}{Friendship Data} & \multicolumn{3}{|c|}{Senate Data} \\
  \hline
  M1 & M2 & M3 & M1 & M2 & M3 \\
  $\alpha_1$ & \Sexpr{SFE$M1[1]} & \Sexpr{SFE$M2[1]} & \Sexpr{SFE$M3[1]} & \Sexpr{SSE$M1[1]} & \Sexpr{SSE$M2[1]} & \Sexpr{SSE$M3[1]} \\
  $\alpha_2$ & \Sexpr{SFE$M1[2]} & \Sexpr{SFE$M2[2]} & \Sexpr{SFE$M3[2]} & \Sexpr{SSE$M1[2]} & \Sexpr{SSE$M2[2]} & \Sexpr{SSE$M3[2]}  \\
  $\alpha_3$ & -- & -- & -- & \Sexpr{SSE$M1[3]} & \Sexpr{SSE$M2[3]} & \Sexpr{SSE$M3[3]} \\
  $\beta_1$ & \Sexpr{SFE$M1[3]} & \Sexpr{SFE$M2[3]} & \Sexpr{SFE$M3[3]} & \Sexpr{SSE$M1[4]} & \Sexpr{SSE$M2[4]} & \Sexpr{SSE$M3[4]} \\
  $\beta_2$ & \Sexpr{SFE$M1[4]} & \Sexpr{SFE$M2[4]} & \Sexpr{SFE$M3[4]} & \Sexpr{SSE$M1[5]} & \Sexpr{SSE$M2[5]} & \Sexpr{SSE$M3[5]}\\
$\beta_3$ & -- & \Sexpr{SFE$M2[5]} & -- & -- & \Sexpr{SSE$M2[6]} & -- \\
  $\beta_4$ & -- & --  & \Sexpr{SFE$M3[6]} & -- & -- & \Sexpr{SSE$M3[7]} \\
  \hline
\end{tabular}
}
\caption{\label{tab:meanests} The means (standard deviations) of parameter values estimated from repeated fittings of $M1, M2, M3$ to the small friendship network and the senate collaboration network. Each model was fit 1,000 times to the friend data, while each model was fit 100 times to the senate data.}
\end{table}

For comparison, we also fit models M1, M2, and M3 to the senate collaboration data. Comparison of the parameter estimates of these models fit to the two different data sets is given in Table~\ref{tab:meanests}. In addition, we fit two more models, M4 and M5 to the senate data that were found to be significant according to the Wald-type test:

@@

In M4, $s_{i5} = \mathbb{I}(x_{i+} = x_{+i} = 0)$ is an indicator variable for isolated nodes, called the network-isolate effect. This parameter attempts to measure the effect of a node being completely isolated in the network. If the value of the parameter $\beta_5$ is high, nodes are encouraged to remain isolated. In M5,  $s_{i6} = \sum\limits_j x_{ij}x_{ji}\mathbb{I}(z_i = z_j)$ is a measure of reciprocity between senators in the same party, $z_i, z_j \in \{\text{Democrat, Independent, Republican}\}$, generally called the same covariate reciprocity effect. This parameter attempts to measure the propensity for forming reciprocal ties between nodes with the same covariate value. If the value of the parameter $\beta_6$ is high, nodes belonging to the same covariate group are encourage to reciprocate incoming ties. By exploring these models more visually, we hope to understand these effects and the model fitting process better.

@@

The first way we hope to better understand stochastic actor-oriented models is by viewing the model(s) in the data space. In \citeauthor{modelvis}, they define the \textit{data space} as ``the region over which we can reliably make inferences, usually a hypercube containing the data" \citep[p.~206]{modelvis}. But what does this definition mean for network data? For dynamic social networks, there are a few different data ``spaces": 
\begin{enumerate}
\item the actors and their corresponding covariates,
\item the edges and their correpsonding variables that describe the relationships between the nodes, and 
\item the time over which the network evolves
\end{enumerate}
These three data pieces can be visualized together in various ways. The traditional node-link visualization uses one of many algorithms to layout the actors as points in 2D space, then draws segments connecting the points in 2D if there is an edge between two nodes, and draws nothing otherwise. The time aspect can be visualized by drawing each network observation in time and placing the observed timepoints side-by-side.

One tool that can bring these three data spaces together in this way is the \texttt{R} package \texttt{geomnet} \citep{geomnet}. Different aesthesic aspects of each piece of the node-link diagram can be tied to the underlying node or edge data. The color, size, and shape of the points can be used to represent variables in the node data, while the color, linewidth, and linetype of the lines between points can be used to represent the edge variables. In a social network, node data might be age, gender, and occupation of the person in the network, and edge data might be length of connection between two people, how the people first met (school, work, church, etc.), and how often they interact, and we can view the network at different timepoints side-by-side to see its evolution. Pulling all of this information together with \texttt{geomnet} allows the entire data space to be viewed at once.

<<funsenatevis, fig.height=3, fig.cap='The 111th Senate separated into ``BC" and ``AC": before Clinton left the Senate to become Secretary of State (on the right), and after she left (on the left).' >>=
se111clint <- filter(seobama, senate == 111)
se111noclint <- filter(seobama, senate == 111)
se111noclint$target[which(se111noclint$target == "Hillary Rodham Clinton")] <- NA
se111clint$Clinton <- 'Yes'
se111noclint$Clinton <- 'No'
clintonSenate <- rbind(se111clint, se111noclint)

ggplot(data = clintonSenate) + 
  geom_net(directed = T, labelon=F, arrowsize = .3, singletons= T, fiteach = T, arrowgap = .01, layout.alg = 'fruchtermanreingold',
           aes(from_id = source, to_id = target, color = party, linewidth = gsw, shape = sex,size = n_au)) + 
  scale_size_continuous(name="# bills\nauthored") +
  #theme_net() + 
  ThemeNet + 
  scale_shape_manual(values = c(17,16)) + 
  scale_color_manual(values = c("royalblue", "forestgreen","firebrick")) + 
  theme(legend.position = 'bottom') + 
  facet_wrap(~Clinton, nrow = 1, labeller = 'label_both')
@

To demonstrate, we use \texttt{geomnet} to visualize the connections in the $111^{th}$ at two different points: when Hillary Clinton was in the senate, and after she left to become Secretary of State. Clinton was only in the $111^{th}$ senate for 17 days, from January 3, 2009, to January 20, 2009, when she was in the middle of her second term in the senate. In that time, she authored two bills and was a cosponsor on 17 other bills. With Clinton included in the node-link diagram, the senate looks much more collaborative than it does without her in the diagram. We can compare the number of bills authored throughout the senate by mapping the size of the node to the numer of bills authored by that senator. We also map shape to gender, and keep color mapped to party of the senator. In addition, we can see the strength of the tie by mapping the linewidth of the edge to the $WPC$ value between the two senators. In this single visualization, we have viewed node information (number of bills authored by a senator and the gender of that senator), edge infortmation (the direction and strength of ties between two senators), and time (before and after Clinton left the senate). 
% ST: I don't like this paragraph
% The visual aspects of the node-link diagram connecting the node, edge, and time data spaces are themselves part of the data space, because they can be viewed in different ways. Another way to connect the node and the edge data, as discussed in Section~\ref{sec:netvizintro}, is through the network's adjaceny matrix. The time dimension can also be incorporated with side-by-side adjacency matrix visualizations.

>>>>

%---------------------------------------------------
%                 Placing Figures
\renewcommand{\topfraction}{0.99}	% max fraction of floats at top
\renewcommand{\bottomfraction}{0.8}	% max fraction of floats at bottom
\renewcommand{\textfraction}{0.05}	% allow minimal text w. figs

@@

\st{
%\par In order to estimate the parameters of  SAOMs the software SIENA, and its \texttt{R} implementation \texttt{RSiena}, was developed \cite{RSiena}. This software marks a huge contribution to the field of social network analysis., and when we began working with it we were eager to understand how it works. HH: we are still interested in this.
%\hh{The next sentence is very vague - make it more specific: what aspect of SAOMs is not well understood? I think what you want is something like ``SAOMs are not very tractable analytically with traditional methods, e.g. likelihood functions turn into very complex objects because of the sheer number of parameters involved. Therefore, computationally more tractable solutions are used to fit estimators, in particular, SAOMs are fit using ... }
Unlike other network models, SAOMs are not very well understood. They are relatively new, especially compared to the classic exponential random graph models, and they are not very tractable analytically. Likelihood functions quickly very complex objects to analyze due to the dependency structure inherent in the data. Therefore, computationally more tractable solutions are used to fit estimators, and in particular, % Traditional modeling steps, like determining the likelihood function, are very complex. Because of this, 
SAOMs are often fit to data using a series of Markov Chain Monte Carlo (MCMC) phases for finding method of moments estimators. In order to estimate the parameters of SAOMs, we use the software SIENA, and its \texttt{R} implementation \texttt{RSiena}, which was developed by \citet{RSiena}. This software marks a huge contribution to the field of social network analysis, but the many moving pieces involved in parameter estimation are largely ``behind the scenes" and hidden from the software user.} In this paper, in order to better understand the model-fitting process, we attempt to bring SAOM fits and the fitting process out of their black boxes, by combining the principals of network visualization with those of model visualization as discussed in \citet{modelvis}. %Here, we explore RSiena and SAOMs with innovative network visualizations, both static and dynamic.
By bringing some light to the underlying methods and structures that are behind the scenes when fitting SAOMs to network data, we aim to help researchers working with the models better understand the implications and analyses of these models.

@@

The \texttt{RSiena} software that fits the models to data also includes a goodnes-of-fit function for examining model fit, \texttt{sienaGOF()}. This function ``assess[es] the fit of the model with respect to auxiliary statistics of networks" \citep[p.~53]{RSienamanual}. Examples of auxiliary statistics include the out- or indegree distribution on the nodes, with the option for users to input their own statistics to examine. The goodness-of-fit is evaluated as follows:
\begin{enumerate}
\item The auxiliary statistics are computed on the observed data and on $N$ simulated observations from the model. Typically, $N=1000$. 
\item The mean vector and covariance matrix of the statistics on the simulations from the model are computed.
\item The Mahalanobis distance from the observed statistics to the distribution of the simulated statistics is computed using the mean and covariance found in step 2. 
\item The Mahalanobis distance from each of the $N$ simulations to the same distribution is computed, and the Mahalanobis distance of the observed data is compared to this distribution of distances.
\item An empirical $p$-value is found by computing the proportion of simulated distances found in step 4 that are as large or larger than the Mahalanobis distance from the data. A SAOM is thus considered a good fit if $p$ is large.
\end{enumerate}

@@

The \texttt{plot.sienaGOF()} function allows us to visualize this fit. %In Figure~\ref{fig:sienagof}, we provide an example of what the goodness-of-fit plot output looks like for indegree and outdegree statistics for a small model. 
This function draws a box plot and a violin plot at each value of the statistic of interest observed in the simulations: \st{on the $x$-axis, the out(in)degrees observed in the data ($0, 1, 2, 3, \dots$), and on the $y$-axis, the cumulative number of times that out(in)degree value appears in the simulations and the data.} In order to compare the distribution of the counts of nodes with the specified degree as calculated on the simulated networks to the counts observed in the true data, red points connected by red lines representing the observed values are superimposed on the boxplots. If the red points lie "well within" the simulated values, the model is a good fit to the data. \st{This plot is not shown because it is not intuitive for understanding network data. Boxplots separated by the many outdegree values observed do not communicate \textit{how} the nodes are connected, just that some have more connections than others. In order to understand the fit of the model, we should try to understand how well the model captures the\textit{overall} structure of the network, not just one or two summary measurements of that structure.} %The model examined in Figure~\ref{fig:sienagof} appears to do a better job of capturing the outdegree distribution than the indegree distribution of the data. 
In Section~\ref{sec:minds}, we propose a new way of visualizing goodness-of-fit that uses the traditional node-link diagram to visualize the entire network instead of numerical summaries of the the network.

@@

<<alldataviz, fig.cap="The smaller friendship network data we will be modelling throughout the paper. In the first wave, we can see there are two large, separate friend group. By the second wave, three students with heavier drinking behavior have separated from their group, while the others become members of the other group. By the third wave, the large group has almost completely broken up. We want to capture these changes with our SAOMs.", out.width='.99\\textwidth', fig.height=3>>=
actual1 <- merge(data.frame(as.edgelist(as.network(fd2.w1))), 
                 data.frame(id = 1:16, drink = drink2[,1]),
                 by.x = "X1", by.y = "id", all = T)
actual1$Wave <- 1 
actual2 <- merge(data.frame(as.edgelist(as.network(fd2.w2))),
                 data.frame(id = 1:16,drink = drink2[,2]),
                 by.x = "X1", by.y = "id", all = T)
actual2$Wave <- 2
actual3 <- merge(data.frame(as.edgelist(as.network(fd2.w3))),
                 data.frame(id = 1:16,drink = drink2[,3]),
                 by.x = "X1", by.y = "id", all = T)
actual3$Wave <- 3
alldat <- rbind(actual1, actual2, actual3)
#alldat$X1 <- paste0("V", alldat$X1)
alldat$X1 <- as.factor(alldat$X1)
#alldat$X2 <- ifelse(!is.na(alldat$X2), paste0("V", alldat$X2), NA)
alldat$X2 <- as.factor(alldat$X2)
#write_csv(alldat, "smallfriends4Geomnet.csv")
alldataviz <- ggplot(data = alldat, 
       aes(from_id = X1, to_id = X2, color = as.ordered(drink), group = as.ordered(drink))) +
  geom_net(fiteach = FALSE, directed=TRUE, layout.alg = 'fruchtermanreingold', size = 5, arrowsize = .25, arrowgap = .05, ecolour = 'grey40', labelon=T, labelcolour = 'black', fontsize = 4, vjust = .5) +
  scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior", labels = c("never", "1-2 times per year", "once per month", "once per week")) + 
  ThemeNet +
  facet_wrap(~Wave, labeller = "label_both") + 
  theme(legend.position = 'bottom', 
        panel.background = element_rect(fill = NA, color = "grey30"), 
        legend.text = element_text(size = 12, family = "Times New Roman"), 
        legend.title = element_text(size = 12, family = "Times New Roman"))
alldataviz
@

@@

Every good data analysis includes both numerical and visual summaries of the data, so why restrict model description and diagnostics to numerical summaries? The concept of model visualization was developed to complement traditional model diagnostic tools. Typically, numerical summaries such as $R^2$ are used to assess model fit, and the occasional visualization, like a residual plot, are used to determine how well the model fits the data. \citeauthor{modelvis} outline three separate ideas, each of which are can be referred to simply as the "model": the model family, the model form, and the fitted model. The latter is primarily what one thinks of first when considering a model in a data analysis, where specified a model is fit to data, and parameter estimates and other numerical summaries, such as $R^2$ are reported. In the context of SAOMs, the fitted model contains the form of the rate and objective functions, the estimated rate parameters, and the estimated objective function parameters. The model form describes the the model \textit{before} the fitting process, defining which parameters are in the model within the context of the larger model family. In SAOMs, the model form includes description of the rate and objective functions and the variables therein that describe how the network evolves over time. Finally, the model family is the broadest description of the model. This is the type of model that you wish to fit to the data, and is chosen based on the problem, data, and knowledge at hand. For example, we chose to use a SAOM to model network data over an exponential random graph model (ERGM) because we believe that actor-level variables effect network structure and formation, and we wanted to model the network changes over time.

%These three uses of the term ``model" 
\st{The model family, the model form, and the fitted model can each be visualized according to the three principals of model visualization: we can }view the model in the data space, visualize collections of models, and explore the process of fitting the model, not just the end result. \st{Since we have already decided on our model family, SAOMs, }we now shift our focus to the fitted model and the model form. Specifically, we want to learn more about how the model form we choose affects the fitted model by using our example data sets and our visualization toolbox. We begin by introducing the five models that we fit to our example data. \st{Next we use the five models to guide our visual explorations of SAOMs. We first use novel tools and ideas to view a SAOM in the data space of a dynamic network. We then explore collections of the same models fit many times to the example data to see how the simulation processes in \texttt{RSiena} affect the model fits. Finally, we look behind the scenes and into the individual steps of the continuous time Markov chain to learn more about how this "hidden data" mechanism works and how it results in a fitted model. }

@@

where $s_{i3} = \sum_{j \neq h} x_{ij}x_{ih}x_{hj} \mathbb{I}(z_i = z_h \neq z_j)$, and $s_{i4} =  |\{j : x_{ij} = 0, \sum\limits_h x_{ih}x_{hj} \geq 2\}|$. These statistics are known as the number of jumping transitive triplets and the number of doubly achieved distances two effect, respectively. The first statistic emphasizes triad relationships that are formed between actors from different covariate groups, while the other emphasizes indirect ties between actors. The covariate groups are determined by the student's drinking behavior, the values of which are numeric and mean-centered. The four values in our data are $\{\text{never, 1-2 times per year, once per month, once per week}\}$, which after being converted to numeric and mean-centered become $\{-0.8125,-1.8125,0.1875,1.1875\}$. The two additional effects are visually represented and further described in Figure~\ref{fig:jtt} and Figure~\ref{fig:dad}, respectively.

@@

In Section~\ref{sec:samemoddiffdat}, we also fit models M1, M2, and M3 to the senate collaboration data for comparison. %Comparison of the parameter estimates of these models fit to the two different data sets is given in Table~\ref{tab:meanests}. 
In addition, we fit two more models, M4 and M5 to the senate data that were found to be significant according to the Wald-type test:

@@

In M4, $s_{i5} = \mathbb{I}(x_{i+} = x_{+i} = 0)$ is an indicator variable for isolated nodes, called the network-isolate effect. This parameter attempts to measure the effect of a node being completely isolated in the network. If the value of the parameter $\beta_5$ is high, nodes are encouraged to remain isolated. In M5,  $s_{i6} = \sum\limits_j x_{ij}x_{ji}\mathbb{I}(z_i = z_j)$ is a measure of reciprocity between senators in the same party, $z_i, z_j \in \{\text{Democrat, Independent, Republican}\}$, generally called the same covariate reciprocity effect. The values of the covariate are converted to numeric, $\{1,2,3\}$, respectively, and mean-centered within \texttt{RSiena}. The mean-centered values of $z \in \{\text{Democrat, Independent, Republican}\}$ are $\{-0.96, 0.04, 1.04\}$. The parameter $\beta_6$ attempts to measure the propensity for forming reciprocal ties between nodes with the same covariate value. If the value of the parameter $\beta_6$ is high, nodes belonging to the same covariate group are encourage to reciprocate incoming ties. \st{Due to the intractability of SAOMs, it is difficult to know for certain how to interpret a fitted value of a parameter. We can make educated guesses based on the definition of the effect and the sign of the fitted value, but a direct interpretation is not always possible. But,} by exploring these models more visually, we aim to understand these effects, their interpretations, and the model fitting process better.

@@

The first way we hope to better understand stochastic actor-oriented models is by viewing the model(s) in the data space. In \citeauthor{modelvis}, they \st{happen to} define the \textit{data space} as ``the region over which we can reliably make inferences, usually a hypercube containing the data" \citep[p.~206]{modelvis}. But what does this definition mean for network data? For dynamic social networks, there are a few different data ``spaces": 
\begin{enumerate}
\item the actors and their corresponding covariates,
\item the edges and their correpsonding variables that describe the relationships between the nodes, and 
\item the time, both the continuous unobserved time and the discrete observed time points, over which the network evolves
\end{enumerate}
These three data pieces can be visualized together in various ways. The traditional node-link visualization uses one of many algorithms to layout the actors as points in 2D space, then draws segments connecting the points in 2D if there is an edge between two nodes, and draws nothing otherwise. The time aspect can be visualized by drawing each network observation in time and placing the observed timepoints side-by-side.

\st{Because longitudinal network data consist of three different ``spaces" of data, viewing the model in the data space can depend on which aspect of the \textit{model} and the \textit{data} we are interested in viewing. %The SIENA software includes a method for modelling covariate change in the network 
Incorporating data covariates into the network structure allows us to assess whether the ties between nodes are affected by how nodes behave over time. In this instance, we would want to view predictions over time. A SAOM can also model behavior change over time, taking both the node and edge information into account. In this case, a plot of predicted covariate values over time would put the model into the time and node data space. Most likely, however, is that we would want to view all of the data spaces simultaneously.}

One tool that can bring the \st{node, edge, and time} data spaces together in this way is the \texttt{R} package \texttt{geomnet} \citep{geomnet}. Different visual features in the node-link diagram can be tied to the underlying node or edge data. The color, size, and shape of the points can be used to represent variables in the node data, while the color, linewidth, and linetype of the lines between points can be used to represent the edge variables. In a social network, node data might be age, gender, and occupation of the person in the network, and edge data might be length of connection between two people, how the people first met (school, work, church, etc.), and how often they interact, and we can view the network at different timepoints side-by-side to see its evolution. Pulling all of this information together with \texttt{geomnet} allows the entire data space to be viewed at once.

<<funsenatevis, fig.height=3, fig.cap='The 111th Senate at two discrete time points: while Clinton was in the senate in the first few weeks of 2009 (on the left) and after she left the senate to become Secretary of State (on the right). We put some potential model covariates, sex and party of the senator and the number of bills they authored in the data space through the shape, color, and size of the nodes, respectively. We also map the strength of the tie, the $WPC$, to the width of the edge between two nodes. Thicker lines implies higher propensity to collaborate. In addition, senators with no ties higher than $WPC = 0.25$ are not shown.'>>=
se111clint <- filter(seobama, senate == 111)
se111noclint <- filter(seobama, senate == 111)
se111noclint$target[which(se111noclint$target == "Hillary Rodham Clinton")] <- NA
se111clint$Clinton <- 'Yes'
se111noclint$Clinton <- 'No'
clintonSenate <- rbind(se111clint, se111noclint)
clintonSenate$Clinton <- as.factor(clintonSenate$Clinton)
clintonSenate$Clinton <- ordered(clintonSenate$Clinton, levels = c("Yes", "No"))

clintonSenate %>% filter(source %in% target)

clintonSenate %>% filter(!(source %in% c("Roland Burris", "Bernard Sanders", "Frank R. Lautenberg", "Judd A. Gregg", "Mary L. Landrieu") & is.na(target))) %>%
ggplot() + 
  geom_net(directed = T, labelon=F, arrowsize = .3, singletons= F, fiteach = T, arrowgap = .01, layout.alg = 'fruchtermanreingold',
           aes(from_id = source, to_id = target, color = party, linewidth = gsw, shape = sex,size = n_au)) + 
  scale_size_continuous(name="# bills\nauthored") +
  #theme_net() + 
  ThemeNet + 
  scale_shape_manual(values = c(17,16)) + 
  scale_color_manual(values = c("royalblue", "forestgreen","firebrick")) + 
  theme(legend.position = 'bottom') + 
  facet_wrap(~Clinton, nrow = 1, labeller = 'label_both')
@

To demonstrate, we use \texttt{geomnet} to visualize the connections in the $111^{th}$ United States Senate at two different points: when Hillary Clinton was in the senate, and after she left to become Secretary of State. Clinton was only in the $111^{th}$ senate for 17 days, from January 3, 2009, to January 20, 2009, when she was in the middle of her second term as senator from New York. In that time, she authored two bills and was a cosponsor on 17 other bills. With Clinton included in the node-link diagram, the senate looks much more highly collaborative than it does without her in the diagram. We can compare the number of bills authored throughout the senate by mapping the size of the node to the numer of bills authored by that senator. We also map shape to sex, and keep color mapped to party of the senator. In addition, we can see the strength of the tie by mapping the linewidth of the edge to the $WPC$ value between the two senators. In this single visualization, we have viewed node information (number of bills authored by a senator and the sex and party of that senator), edge information (the direction and strength of ties between two senators), and time (before and after Clinton left the senate). 
% ST: I don't like this paragraph
% The visual aspects of the node-link diagram connecting the node, edge, and time data spaces are themselves part of the data space, because they can be viewed in different ways. Another way to connect the node and the edge data, as discussed in Section~\ref{sec:netvizintro}, is through the network's adjaceny matrix. The time dimension can also be incorporated with side-by-side adjacency matrix visualizations.

% below chunk contains some repetitive stuff. Haven't figured out what it's purpose is yet. 
<<summnetM1, message=FALSE, fig.cap='On the left, the first wave of observed data that is conditioned on in the model. On the right, the second wave of observed data. In the middle, a summary network from the first model fit to the data. This summary network represents 1,000 simulations of wave 2 using the values from the simple fitted model M1.',results='hide', fig.height=2.5>>=
library(netvizinf)
friendshipData <- array(c(fd2.w1, fd2.w2, fd2.w3), dim = c(16, 16, 3))
drink1 <- s50a[20:35,1]
drink2 <- s50a[20:35,2]
drink3 <- s50a[20:35,3]
friendship <- sienaDependent(friendshipData)
alcohol <- varCovar(cbind(drink1,drink2, drink3))
mydata <- sienaDataCreate(friendship, alcohol)
M1eff <- getEffects(mydata)
M2eff <- includeEffects(M1eff, jumpXTransTrip, interaction1 = "alcohol")
M3eff <- includeEffects(M1eff, nbrDist2twice)
simu2 <- read.csv("data/simulation-1000-M1-M2-M3.csv")
means <- simu2 %>% group_by(Model, parameter) %>% summarize(
  means = mean(estimate)
)
M1parms <- (means %>% filter(Model == "M1"))$means
M2parms <- (means %>% filter(Model == "M2"))$means
M3parms <- (means %>% filter(Model == "M3"))$means
set.seed(4231352)
M1sims1000 <- netvizinf::saom_simulate(dat = mydata, 
                            struct = M1eff,
                       parms = as.numeric(M1parms), 
                       N = 1000)
M2sims1000 <- netvizinf::saom_simulate(dat = mydata, 
                            struct = M2eff,
                       parms = as.numeric(M2parms), 
                       N = 1000)
M3sims1000 <- netvizinf::saom_simulate(dat = mydata, 
                            struct = M3eff,
                       parms = as.numeric(M3parms), 
                       N = 1000)
M1simsdf <- sims_to_df(M1sims1000)
M2simsdf <- sims_to_df(M2sims1000)
M3simsdf <- sims_to_df(M3sims1000)
M1avgW2 <- M1simsdf %>% filter(!is.na(to) & wave == 1) %>%
  group_by(from, to) %>% 
  summarise(count = n()) %>% 
  mutate(weight = ifelse(from == to, 0, count))
M2avgW2 <- M2simsdf %>% filter(!is.na(to) & wave == 1) %>%
  group_by(from, to) %>% 
  summarise(count = n()) %>% 
  mutate(weight = ifelse(from == to, 0, count))
M3avgW2 <- M3simsdf %>% filter(!is.na(to) & wave == 1) %>%
  group_by(from, to) %>% 
  summarise(count = n()) %>% 
  mutate(weight = ifelse(from == to, 0, count))
